Financialization of Commodities: an Asset Pricing Perspective

### 2024-06-11 others/paper/custom/anonymous.tex
Track 'others' folder

\documentclass[11pt, letterpaper, doublespacing]{article}

\usepackage{setspace}

\makeatletter

\renewcommand{\@seccntformat}[1]{\@nameuse{the#1}.~{}}

\renewcommand{\thesection}{\Roman{section}}
\renewcommand{\thesubsection}{\Alph{subsection}}
\renewcommand{\thesubsubsection}{\arabic{subsubsection}}

\renewcommand{\section}{\@startsection {section}{1}{\z@}%
  {-3.5ex \@plus -1ex \@minus -.2ex}%
  {2.3ex \@plus.2ex}%
  {\centering\normalfont\large\bfseries}}

\renewcommand{\subsection}{\@startsection{subsection}{2}{\z@}%
  {-3.25ex\@plus -1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex}%
  {\normalfont\large\slshape}}

\renewcommand{\subsubsection}{\@startsection{subsubsection}{3}{\z@}%
  {-3.25ex\@plus -1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex}%
  {\centering\normalfont\slshape}}

\makeatother

% \newtheoremstyle{plain}{9pt}{9pt}{\itshape}{0pt}{\scshape}{:}{3pt}{}

% \theoremstyle{plain}

% \newtheorem{assumption}{Assumption}
% \newtheorem{corollary}{Corollary}
% \newtheorem{definition}{Definition}
% \newtheorem{lemma}{Lemma}  
% \newtheorem{proposition}{Proposition}
% \newtheorem{remark}{Remark}
% \newtheorem{theorem}{Theorem}

% \usepackage[bf,small,nooneline,normal]{caption}
% \renewcommand*{\captionlabeldelim}{\nobreak}
% \renewcommand*\captionlabeldelim{.}
% \renewcommand{\captionfont}{\bfseries}

\renewcommand{\thetable}{\Roman{table}}

\usepackage{indentfirst}

\renewcommand{\abstractname}{\bf \small ABSTRACT}

\renewcommand{\refname}{\bf \small REFERENCES}

\usepackage{endnotes}
\renewcommand{\footnote}{\endnote}
\renewcommand{\notesname}{\bf Footnotes}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Financialization of commodities: an asset pricing perspective},
            pdfauthor={Devraj Basu, Olivier Bautheac},
            pdfkeywords={Financialization of commodities, futures contracts in agricultural
commodities, speculative bubbles, volume, open interest, convenience
yield.},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{jf}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Financialization of commodities: an asset pricing perspective}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Devraj Basu, Olivier Bautheac}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{caption}
\usepackage{setspace}
\doublespacing

\begin{document}
\maketitle
\begin{abstract}
The last two decades have seen major changes in the nature of the
commodity markets with the influx of financial investor inflows followed
by the financial crisis. We study the effect of these events on
commodity futures pricing dynamics using commodity asset pricing
techniques. We consider both futures market based factors and the
traditional market factor and find that a hedging pressure based factor
is able to best detect the effects of commodity financialization while
the effects of the financial crisis and after are most clearly visible
through the performance of the market factor. The futures market seems
to have been the main channel for the effects of commodity
financialization, which seems to have delivered an exogenous schock to
the cross-section of global commodities.
\end{abstract}

@@

\section{Introduction}\label{introduction}

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialization\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialization of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialization appears to have
affected a number of economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A vigorous legislative debate has led
to regulatory changes\footnote{Concerns over the consequences of
  financialization were behind Rule 76 FR 4752 issued by the U.S.
  Commodity Futures Trading Commission (CFTC) on January 26, 2011. This
  provision emanates from the Dodd-Frank Wall Street and Consumer
  Protection Act of 2010 (Title VII, Section 737) that mandates the CFTC
  to use position limits to restrict the flow of speculative capital
  into a number of commodity markets. The Rule was approved in a close
  3-2 vote and the ensuing rulemaking process was extremely contentious
  with several commissioners (Michael Dunn and Scott D. O'Malia in
  particular) expressing reservations about the lack of supporting
  evidence and the Rule also triggering thousands of comment letters as
  well as a lawsuit against the CFTC spearheaded by two Wall Street
  trade groups, namely the International Swaps \& Derivatives
  Association (ISDA) and the Securities Industries \& Financial Markets
  Association (SIFMA) (ISDA \& SIFMA v U.S. CFTC; Complaint,
  1:11-cv-02146; December 2, 2011).} accompanied by an equally intense
policy and academic debate across a number of disciplines about the
perceived effects on commodity prices of the unprecedented inflow of
institutional funds brought into commodity futures markets by new index
type investors\footnote{The first responses to the 2007/2008 crisis of
  escalating food and energy prices took the form of policy reports,
  many of which reasoned that the growth of commodity index funds came
  along with an influx of largely speculative capital that was
  responsible for driving commodity prices beyond their historic highs
  \citep{de_schutter_food_2010_1, gilbert_speculative_2010, herman_not_2011_1, schumann_hunger-makers_2011, unctad_global_2009}.}.
In response to immediate policy concerns, the academic debate was
initially framed rather narrowly\footnote{Early U.S. Senate
  investigations drew on pricing and trading data supplied by the CFTC
  as well as interviews with numerous experts including hedge fund
  manager Michael Masters who linked the growing presence of index
  investors in U.S. wheat markets to the observed surges in both futures
  and cash prices
  \citep{masters_testimony_2008, masters_accidental_2008}. This
  contention has since come to be known as the Masters' hypothesis and
  was largely endorsed by the final U.S. Senate report on the issue
  \citep{USSenate_Excessive_speculation_wheat_2009}. At this stage, the
  question thrust on the academic sphere was whether ``excessive
  speculation'' (understood as the peculiar market activity of index
  type investors) was linked to escalating energy and agricultural
  commodity prices.} and used a fairly restricted set of statistical
techniques\footnote{Most of the studies focused on statistical analyses
  of the Granger type causal connections between ``excessive
  speculation'' and commodity prices.}.

In this paper we take a fundamentally different empirical approach that
is broader in scope. We consider the entire cross-section of actively
traded commodities on futures markets and use a futures based asset
pricing framework that includes factors constructed using both price and
non-price attributes. Factor model techniques are well suited to
isolating common driving factors and detecting co-movement, a central
issue in financialization which we believe the existing literature has
not fully analyzed.\\
The first factor model we consider (CHP) is based on the well-known
hedging pressure based theory
\citep{anderson_hedger_1983, chang_returns_1985, cootner_returns_1960, dusak_futures_1973, hicks_value_1939, hirshleifer_risk_1988, hirshleifer_determinants_1989, hirshleifer_hedging_1990, kolb_is_1992, keynes_treatise_1930}
which postulates that futures prices for a given commodity are inversely
related to the extent that commercial hedgers are short or long and we
use a factor mimicking portfolio that captures the impact of hedging
pressure as a systemic factor \citep{basu_capturing_2013}. Our second
factor model is based on open interest growth following
\citet{hong_what_2012} who find that open interest growth has predictive
power for individual commodity returns and again we construct a
long-short factor mimicking portfolio. Our third model is based on term
structure following \citet{szymanowska_anatomy_2014} and
\citet{fuertes_commodity_2015} who demonstrate that the term structure
factor has explanatory power for the cross-section of commodity returns.
We finally consider an equally weighted portfolio of commodities as a
proxy for the market factor. Our two price based and two non-price based
factors allow us to compare and contrast different channels that could
have led to changes in commodity price dynamics over the last twenty
years.

We study the performance of these models on a set of twenty-four U.S.
traded commodities and six non-U.S. traded (LME traded) industrial
metals over the 1997/2017 period that we divide into four distinct
sub-periods. The first, the pre-financialization phase, spans the
1997/2003 period; the second, the financialization phase, spans the
2004/2007 period\footnote{Starting point based on earlier studies
  \citep{baker_financialization_2014, christoffersen_factor_2014}.}; the
third, 2008/2013, encompasses both the financial crisis and the loose
monetary policy regimes that followed; and the fourth spans the period
starting in 2014 to present.\\
Consistent with earlier studies, we observe that mean returns for the
U.S. traded commodities are sharply higher over the financialization
period compared to pre-financialization and this pattern extends to LME
traded metals.

The cross-sectional nature of financialization
\citep{basak_model_2016, cheng_financialization_2014} leads us to
extends the Keynesian indivudal hedging pressure paradigm by
incorporating aggregate CHP (\(\mathbf{CHP}\)), a market wide measure of
hedging pressure. This paradigm implies that returns for individual
commodities should be high in periods where \(\mathbf{CHP}\) is low
(\(\mathbf{backwardation}\)) and low in periods where \(\mathbf{CHP}\)
is high (\(\mathbf{contango}\)) vice et versa. Over the first period the
cross sectional average returns during \(\mathbf{backwardation}\) for
both the US and UK commodities are positive with those for the US
statistically significant, and significantly higher than during
\(\mathbf{contango}\), providing broad support for an aggregate version
of the Keynesian hedging pressure hypothesis. Over the second period
this pattern reverses, with both the US and UK commodities having
significant positive returns during \(\mathbf{contango}\) with these
returns being significantly higher than during
\(\mathbf{backwardation}\). This change in pattern suggests that the
onset of financialization altered the traditional risk shifting nature
of hedger's behaviour, which is implied by the Keynesian hypothesis.

We analyze the pricing dynamics over the different periods by first
studying the time series pricing performance of the various factors on
all the U.S. traded commodities with the factors constructed using the
latter set of commodities. The market based factor outperforms the other
three in the first period and it has a similar level of performance in
the second period, while the CHP factor shows the greatest improvement
in performance in the second period. The performance of the market
factor is driven by agricultural commodities and metals in both the
first and second period and the dramatic improvement of the CHP factor
in the second period is driven by precious and precious industrial
metals. The best performing commodities for the market factor exhibited
approximately the same level of price co-movement both before and during
financialization. In contrast, the best performing commodities for the
CHP factor exhibited very low pair-wise correlation in the first period,
around the average for U.S. traded commodities, while during
financialization the average pairwise correlation between the top
performing CHP factor commodities is higher than that for the market
factor and double of the average for all the U.S. traded commodities.
This increase is due to the dramatic rise in average pairwise
correlation amongst the U.S. traded metals all of which exhibited high
R\textsuperscript{2} with respect to the CHP factor during
financialization which in turn is due to their common low level of CHP
over this period. Thus the CHP factor model is able to establish the
strengthening link between a non-price attribute and price dynamics in
the U.S. metal sub-sector during financialization.\\
We explore this issue further by constructing the four factors with the
top eight commodities for a given factor (factor picks) in each
period\footnote{See section 2.2 for further details.} and evaluating its
performance on these selected commodities as well as on all U.S. traded
commodities. The logic of this exercise is to determine the extent to
which the pricing performance of these factors in each period is
concentrated in a subset of commodities. The CHP and the market factors
built from CHP factor picks exhibit the most dramatic increase in
pricing performance on the corresponding factor picks in the second
period. In contrast, the increase in pricing performance on all the U.S.
traded commodities is roughly uniform across all sets of factor models
and lower in magnitude. This strongly suggests that the hedging pressure
co-movement and price co-movement link was concentrated in the metals
sub-sector. There is also a dramatic improvement in pricing performance
of the CHP factor built with term structure picks indicating convergence
between Keynesian and Working's theories of the futures term structure
in a U.S. context.

The LME traded industrial metals provide a useful set of assets to test
how widespread were the changes in pricing dynamics brought about by the
onset of financialization. To that end, we examine the performance of
the four factors on the six LME metals, when constructed from each set
of factor picks. The performance of the market factor built with the CHP
and term structure factor picks improves dramatically in the second
period relative to the first while the performance of all factors built
with the other sets of picks does not increase quite as much. The
pricing performance of the market factor built with CHP factor picks as
well as that of the CHP factor built with both CHP and term structure
factor picks show a dramatic increase over the financialization period.
The performance of the market factor built with the CHP factor picks
suggests that there was co-movement in global metals returns in the
second period and that this co-movement could be detected by our hedging
pressure based model while the performance of the CHP factor built with
term structure factor picks indicates the aforementioned convergence
between Keynesian and Working's views of commodity futures term
structure extends to a global context.

Our analysis suggests that financialization appears to have had the
deepest impact on global metals markets and our commodity asset pricing
based approach allows us to better understand the channels through which
it was transmitted. The initial view of financialization was that it
consisted of speculative flows which had the effect to driving up and
creating bubble like conditions in commodity future prices\footnote{\citet{de_schutter_food_2010_1};
  \citet{gilbert_how_2010}; \citet{gilbert_speculative_2010};
  \citet{masters_testimony_2008}; \citet{masters_accidental_2008};
  \citet{herman_not_2011_1}; \citet{schumann_hunger-makers_2011};
  \citet{singleton_investor_2013}; \citet{unctad_global_2009}}. A number
of academic studies, most of which used commodity specific causality and
correlation based analysis, have disproved this contention, particularly
in the case of agricultural and energy commodities\footnote{\citet{irwin_devil_2009},
  \citet{sanders_adequacy_2010}, \citet{sanders_impact_2011},
  \citet{sanders_new_2011}, \citet{irwin_commodity_2013},
  \citet{brunetti_commodity_2014}, \citet{hamilton_effects_2015} and
  \citet{bruno_financialization_2017} study the agricultural markets.
  \citet{buyuksahin_speculators_2011}, \citet{tokic_speculation_2012},
  \citet{fattouh_role_2013}, \citet{kilian_role_2014},
  \citet{knittel_simple_2016} and \citet{manera_modelling_2016} study
  the energy markets. \citet{bohl_does_2013_2}, \citet{kim_does_2015}
  and \citet{boyd_prevalence_2016} study both energy and agricultural
  markets while \citet{irwin_index_2011},
  \citet{irwin_financialization_2012} and \citet{stoll_commodity_2011}
  study the commodity markets in general.}, while other studies have
underlined various alterations in commodity pricing dynamics over the
period\footnote{\citet{gilbert_role_2014} reject the view that
  financialization has not had any effects in the grains markets and
  demonstrate that trades originated by financial actors, and
  specifically index investors, can move prices but tend typically to be
  volatility-reducing. \citet{cheng_financialization_2014} show that in
  the case of index commodities, financialization has transformed the
  risk sharing, and information discovery functions of commodity futures
  markets. \citet{juvenal_speculation_2015} contend that the oil price
  increase between 2004 and 2008 was mainly driven by the strength of
  global demand but that the financialization process of commodity
  markets also played a role. \citet{henderson_new_2015} show that
  non-information-based financial investments have important impacts on
  commodity prices.}. However, the fundamental question of the nature of
the impact of financialization on commodity futures markets has not yet
been completely answered. Our analysis takes a broader perspective and
shifts the focus away from the energy and agricultural sectors. We
provide an empirical complement to a new stream of theoretical studies
that try to model the impact of financialization on various aspects of
commodity futures markets\footnote{\citet{etula_broker-dealer_2013},
  \citet{acharya_limits_2013}, \citet{cheng_convective_2014},
  \citet{leclercq_equilibrium_2014}, \citet{sockin_informational_2015},
  \citet{goldstein_speculation_2013},
  \citet{EkelandSpeculationcommodityfutures2016},
  \citet{goldstein_commodity_2017}.}, as well as extend the empirical
findings of \citet{basak_model_2016} and \citet{tang_index_2012}. In
contrast to the earlier purely statistical studies that were designed to
study speculative behavior alone, our empirical approach allows us to
isolate the more subtle effects of financialization identified in the
theoretical studies above.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[letterpaper, left = 1in, right = 1in, top = 1.5in, bottom = 1.5in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Financialization of Commodities: an Asset Pricing Perspective},
            pdfkeywords={Commodity futures; financialization; hedging pressure; financial crisis},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{jf}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Financialization of Commodities: an Asset Pricing Perspective}
    \author{}
    \date{}

\usepackage{longtable, tabu, pdflscape, booktabs, caption, setspace}
\doublespacing
\renewcommand*{\arraystretch}{0.65}

\begin{document}
\maketitle
\begin{abstract}
The last two decades have seen major changes in the nature of the
commodity markets with the influx of financial investor inflows followed
by the financial crisis. We study the effect of these events on
commodity futures pricing dynamics using commodity asset pricing
techniques. We find that a non-price based factor is best able to detect
the effects of commodity financialization, while the effects of the
financial crisis and after are most clearly visible through the
performance of the market factor. The effects of financialization seem
to go beyond mechanical effects induced by the rise of indexing. 
\newline
JEL: G11, G23, G13, Q02
\newline
Keywords: Commodity futures, financialization, hedging pressure, financial crisis

\end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialization\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialization of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialization appears to have
affected a number of economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A vigorous legislative debate has led
to regulatory changes accompanied by an equally intense policy and
academic debate across a number of disciplines about the perceived
effects on commodity prices of the unprecedented inflow of institutional
funds brought into commodity futures markets by new index type
investors\footnote{See section \ref{background} for a detailled
  discussion.}. In response to immediate policy concerns the academic
debate was initially framed rather narrowly. In this paper we take a
fundamentally different empirical approach that is broader in scope. We
consider the entire cross-section of actively traded commodities on
futures markets and use a futures based asset pricing framework that
includes factors constructed using both price and non-price attributes.

Commodity futures markets have a history of being the classic physical
commodity risks shifting conduit for commodity specialists, and
financialization may have altered this long standing paradigm with the
entry of long only commodity index traders (CITs) with very different
hedging and volume demands from traditional participants. This change
could have altered the hedging ontology in these markets and our asset
pricing framework is well suited to detect these more subtle effects
driven by changes in the nature of traders. Factor models are well
suited to isolating common driving factors and detecting co-movement and
building on theory we develop a bespoke commodity-based asset pricing
framework that seems well adapted to study these two issues
simultaneously.

The first factor model we consider (CHP) is based on the well-known
hedging pressure based theory
\citep{anderson_hedger_1983, chang_returns_1985, cootner_returns_1960, dusak_futures_1973, hicks_value_1939, hirshleifer_risk_1988, hirshleifer_determinants_1989, hirshleifer_hedging_1990, kolb_is_1992, keynes_treatise_1930}
which postulates that futures prices for a given commodity are inversely
related to the extent that commercial hedgers are short or long and we
use a factor mimicking portfolio that captures the impact of hedging
pressure as a systemic factor \citep{basu_capturing_2013}. Our second
factor model is based on open interest growth following
\citet{hong_what_2012} who find that open interest growth has predictive
power for individual commodity returns and again we construct a
long-short factor mimicking portfolio. Our third model is based on term
structure following \citet{szymanowska_anatomy_2014} and
\citet{fuertes_commodity_2015} who demonstrate that the term structure
factor has explanatory power for the cross-section of commodity returns.
We finally consider an equally weighted portfolio of commodities as a
proxy for the market factor. Our two price based and two non-price based
factors allow us to compare and contrast different channels that could
have led to changes in commodity price dynamics over the last twenty
years.

We study the performance of these models on a set of twenty-four US
traded commodities and six UK traded (LME exchange) base metals over the
1997/2017 period that we divide into four distinct sub-periods. The
first, the pre-financialization phase, spans the 1997/2003 period; the
second, the financialization phase, spans the 2004/2007 period\footnote{Starting
  point based on earlier studies
  \citep{baker_financialization_2014, christoffersen_factor_2018}.}; the
third, 2008/2013, encompasses both the financial crisis and the loose
monetary policy regimes that followed; and the fourth spans the period
starting in 2014 to present.\\
Consistent with earlier studies, we observe that mean returns for the US
traded commodities are sharply higher over the financialization period
compared to pre-financialization and this pattern extends to UK traded
metals.

The arrival of financial investors raises the question of whether they
functioned as liquidity providers or liquidity demanders. Arguments have
been made for both sides with \citet{moskowitz_time_2012} arguing that
financial investors provide liquidity while
\citet{Kangtaletwopremiums2017} argue that they demand liquidity with
\citet{cheng_convective_2014} showing that they in fact transferred risk
to hedgers during the 2008 financial crisis. We study this issue by
extending the Keynesian individual hedging pressure paradigm by
incorporating aggregate CHP (\textbf{CHP}), a market wide measure of
hedging pressure first considered in \citet{hong_what_2012}. Our use of
this measure is motivated by the cross sectional nature of
financialization
\citep{basak_model_2016, cheng_convective_2014, le_futures_2017}. This
paradigm implies that returns for individual commodities should be high
in periods where \textbf{CHP} is low (\textbf{backwardation}) and low in
periods where \textbf{CHP} is high (\textbf{contango}). Over the first
period the cross sectional average returns during \textbf{backwardation}
for both the US and UK commodities are positive with those for the US
statistically significant, and significantly higher than during
\textbf{contango}, providing broad support for an aggregate version of
the Keynesian hedging pressure hypothesis. Over the second period this
pattern reverses, with both the US and UK commodities having significant
positive returns during \textbf{contango} with these returns being
significantly higher than during \textbf{backwardation}. This change in
pattern suggests that the onset of financialization altered the
traditional risk shifting nature of hedger's behaviour implied by the
Keynesian hypothesis and suggests that financial investors may have been
liquidity demanders during financialization leading hedgers to become
liquidity providers.

\citet{goldstein_commodity_2017} as well as
\citet{goldstein_speculation_2014} argue that the arrival of financial
investors has an impact on the extent of normal backwardation in
commodity futures markets. Based on their analysis we would expect our
average CHP factor, which is based on backwardation measured by hedger's
positions, as well our term structure factor, which is based on
backwardation measured by roll yield, to be best suited to measure the
impact of financialization.\\
We analyze the pricing dynamics over the different periods by first
studying the time series pricing performance of the various factors on
all the US traded commodities with the factors constructed using the
latter set of commodities. The market based factor outperforms the other
three in the first period while the CHP factor shows the greatest
improvement in performance in the second period with a four-fold
increase in pricing performance. The performance of the market factor is
driven by agricultural commodities and metals in both the first and
second period and the dramatic improvement of the CHP factor in the
second period is driven by precious and precious-base metals. The best
performing commodities for the market factor exhibited approximately the
same level of price co-movement both before and during financialization.
In contrast, the best performing commodities for the CHP factor
exhibited very low pair-wise correlation in the first period, around the
average for US traded commodities, while during financialization the
average pairwise correlation between the top performing CHP factor
commodities is higher than that for the market factor and double of the
average for all the US traded commodities. This increase is due to the
dramatic rise in average pairwise correlation amongst the US traded
metals all of which exhibited high R\textsuperscript{2} with respect to
the CHP factor during financialization which in turn is due to their
common low level of CHP over this period. Thus the CHP factor model is
able to establish the strengthening link between a non-price attribute
and price dynamics in the US metal sub-sector during financialization.\\
We explore this issue further by constructing the four factors with the
top eight commodities for a given factor (factor picks) in each
period\footnote{See section \ref{methods} for further details.} and
evaluating their performance on these selected commodities as well as on
all US traded commodities. The results of this exercise strongly suggest
that the hedging pressure co-movement and price co-movement link was
concentrated in the metals sub-sector.

The UK traded base metals provide a useful set of assets to test how
widespread were the changes in pricing dynamics brought about by the
onset of financialization. To that end, we examine the performance of
the four factors on the six UK metals, when constructed from each set of
factor picks. The results show that there was co-movement in global
metals returns in the second period and that this co-movement could be
detected by our hedging pressure based model.

The onset of the financial crisis and the monetary policy regimes that
followed, on the other hand, also appear to have induced significant
changes in pricing dynamics. Returns on both US and UK traded
commodities fell dramatically over both the third and fourth periods
with the pattern of return for the US traded commodities across
\textbf{CHP} regimes reverting to the Keynesian paradigm. The pricing
results indicate the emergence of a systematic factor across the entire
cross-section of the US commodities as well as strong evidence of
cross-market linkages.

Commodity financialization raised societal concerns, leading to
legislative action intended to alter the nature and functioning of these
markets. The debate was initially framed around adequacy of speculation,
the burning issue of the day, and the consequent analysis focused on the
more mechanical effects of financialization. With the benefit of
hindsight, it appears this approach was perhaps too narrow and it now
seems necessary to address financialization from a broader perspective.
Commodity price dynamics appear to have altered substantially in quite
different ways over the financialization and the financial crisis
periods and our commodity futures based asset pricing approach seems
able to provide new insights into the nature of these changes:
financialization was a phenomenon endogenous to the commodity markets
transmitted via the commodity futures markets and mainly concentrated in
a specific sector (metals) while the crisis and its aftermath seems to
have delivered an exogenous shock across the entire cross-section of
global commodities. We provide an empirical complement to a new stream
of theoretical studies that try to model the impact of financialization
on various aspects of commodity futures markets\footnote{\citet{etula_broker-dealer_2013},
  \citet{acharya_limits_2013}, \citet{cheng_convective_2014},
  \citet{leclercq_equilibrium_2014}, \citet{sockin_informational_2015},
  \citet{goldstein_speculation_2014},
  \citet{EkelandSpeculationcommodityfutures2016},
  \citet{goldstein_commodity_2017} as well as \citet{le_futures_2017},
  one of the few empirical studies on the impact of financialization on
  commodity-price cross-moments.} and demonstrate that the effects of
financialization extend beyond the mechanical effects induced by
indexation as outlined in \citet{basak_model_2016} as well as
\citet{tang_index_2012}.

\hypertarget{data-methods}{%
\section{Data \& methods}\label{data-methods}}

\hypertarget{background}{%
\subsection{Background}\label{background}}

The period from 1998 to 2008 saw an influx of investment into commodity
index linked products most of which made its way into the commodity
futures market. Exposure to commodity indexes whose goal was to track
the broad movement of commodity prices could be accessed by means of
exchange traded funds and exchange traded notes and at least \$100
billion of new investment flowed into these products over the 1998-2008
period with this financialization phenomenon leading to substantial
legislative changes. Concerns over the consequences of financialization
were behind Rule 76 FR 4752 issued by the US Commodity Futures Trading
Commission (CFTC) on January 26\textsuperscript{th}, 2011. This
provision emanates from the Dodd-Frank Wall Street and Consumer
Protection Act of 2010 (Title VII, Section 737) that mandates the CFTC
to use position limits to restrict the flow of speculative capital into
a number of commodity markets. The Rule was approved in a close 3-2 vote
and the ensuing rulemaking process was extremely contentious with
several commissioners (Michael Dunn and Scott D. O'Malia in particular)
expressing reservations about the lack of supporting evidence and the
Rule also triggering thousands of comment letters as well as a lawsuit
against the CFTC spearheaded by two Wall Street trade groups, the
International Swaps \& Derivatives Association (ISDA) and the Securities
Industries \& Financial Markets Association (SIFMA)\footnote{ISDA \&
  SIFMA v US CFTC; Complaint, 1:11-cv-02146; December
  2\textsuperscript{nd}, 2011.}.

A world-wide debate has ensued about the role of index funds in
commodity markets. The first responses to the 2007/2008 crisis of
escalating food and energy prices took the form of policy reports, many
of which reasoned that the growth of commodity index funds came along
with an influx of largely speculative capital that was responsible for
driving commodity prices beyond their historic highs
\citep{de_schutter_food_2010_1, gilbert_speculative_2010, herman_not_2011_1, schumann_hunger-makers_2011, unctad_global_2009}.
The US and various european countries also became involved in this
issue. Early US Senate investigations drew on pricing and trading data
supplied by the CFTC as well as interviews with numerous experts
including hedge fund manager Michael Masters who linked the growing
presence of index investors in US wheat markets to the observed surges
in both futures and cash prices
\citep{masters_testimony_2008, masters_accidental_2008}. This contention
has since come to be known as the Masters' hypothesis and was largely
endorsed by the final US Senate report on the issue
\citep{USSenate_Excessive_speculation_wheat_2009}. At this stage, the
question thrust on the academic sphere was whether ``excessive
speculation'' (understood as the market activity peculiar to index type
investors) was linked to escalating energy and agricultural commodity
prices. A number of academic studies, most of which used commodity
specific causality and correlation based analysis have disproved this
contention, particularly in the case of agricultural and energy
commodities. \citet{irwin_devil_2009}, \citet{sanders_adequacy_2010},
\citet{sanders_impact_2011}, \citet{sanders_new_2011},
\citet{irwin_commodity_2013}, \citet{brunetti_commodity_2014},
\citet{hamilton_effects_2015} and \citet{bruno_financialization_2017}
study the agricultural markets. \citet{buyuksahin_speculators_2011},
\citet{tokic_speculation_2012}, \citet{fattouh_role_2013},
\citet{kilian_role_2014}, \citet{knittel_simple_2016} and
\citet{manera_modelling_2016} study the energy markets.
\citet{bohl_does_2013_2}, \citet{kim_does_2015} and
\citet{boyd_prevalence_2016} study both energy and agricultural markets
while \citet{irwin_index_2011}, \citet{irwin_financialization_2012} and
\citet{stoll_commodity_2011} study the commodity markets in general.

Other studies have underlined various alterations in commodity pricing
dynamics over the period. \citet{gilbert_role_2014} reject the view that
financialization has not had any effects in the grains markets and
demonstrate that trades originated by financial actors, and specifically
index investors, can move prices but tend typically to be
volatility-reducing. \citet{cheng_financialization_2014} show that in
the case of index commodities, financialization has transformed the risk
sharing, and information discovery functions of commodity futures
markets. \citet{juvenal_speculation_2015} contend that the oil price
increase between 2004 and 2008 was mainly driven by the strength of
global demand but that the financialization process of commodity markets
also played a role. \citet{henderson_new_2015} show that
non-information-based financial investments have important impacts on
commodity prices.

\hypertarget{data}{%
\subsection{Data}\label{data}}
